Local Search Approximation Algorithms for Partitional and Hierarchical Clustering

### 2015-04-06 tex/project.tex
More background and updated bib

If we relax the constraint to have exactly $k$ clusters, then a recent algorithm called affinity propogation has shown lower error and computation time than the algorithms above \cite{Frey}. Instead of specifying a number of clusters, the user inputs ``preferences'' on the data points to be cluster centers. This algorithm relies on passing messages between data points. For each pair of data points $i, k \in X$, ``availability'' $a(i,k)$ represents the evidence for having $i$ choose $k$ to be it's cluster center and ``responsibility'' $r(i,k)$ reflects how well-suited $k$ is to serve $i$. These are messages are then updated iteratively. Once the messages converge (this is not guaranteed) then $\max_{k \in X} a(i,k) + r(i,k)$ will identify the cluster center for $i$, and these decisions will be consistent \cite{Frey}.

@@

While there exist many other methods for clustering such as spectral clustering and mixture models \cite{Murphy}, we will restrict the comparison of our proposed algorithm to the center-based algorithms above which are widely used and relate in objective function.

@@

There are three choices of criteria for choosing which clusters to merge, which reflect the relative benefits of $k$-means, $k$-median, and $k$-center. 
In single linkage clustering (also called nearest neighbor), we measure the distance between two clusters to be the smallest pairwise distance and then merge the two closest clusters. The main potential drawback to this method is that two points within a cluster can be very far apart. Complete linkage clustering on the other hand measures the distance between two clusters to be the largest pairwise distance and then merges the two closest clusters. While points within a single cluster will be quite similar using this metric, points may not belong to their closest cluster. 
Lastly, there is group average linkage which looks at the average pairwise distance between clusters and merges the two closest clusters. This strikes a balance between the closeness and compactness of the two techniques above and gives more reasonable results in practice \cite{ESL}.

@@

Again, several algorithms have been built upon agglomerative clustering to incorporate non-spherical cluster shapes and categorical data such as the CURE, ROCK, and Chameleon algorithms \cite{Guha_CURE, Guha_ROCK, Karypis} but do not relate as much to our proposed algorithm in objective.

@@

>>>>

More recently, spectral based methods have become popular for clustering. These algorithms construct a graph based on the similarity matrix and compute the eigenvalues of the normalized graph Laplacian matrix \cite{Luxburg}. These eigenvectors are either used to split the current cluster in half using the second eigenvector (continuing recursively until there are $k$ clusters) or the eigenvectors are used to project the data points into a lower dimensional space on which the $k$-means algorithm above is run \cite{Verma}. We will base our comparisons against the algorithm introduced by \cite{Meila} which uses the latter method and performed best in practice on several datasets \cite{Verma}.

If we relax the constraint to have exactly $k$ clusters, then a recent algorithm called affinity propagation has shown lower error and computation time than the algorithms above \cite{Frey}. Instead of specifying a number of clusters, the user inputs ``preferences'' on the data points to be cluster centers. This algorithm relies on passing messages between data points. For each pair of data points $i, k \in X$, ``availability'' $a(i,k)$ represents the evidence for having $i$ choose $k$ to be it's cluster center and ``responsibility'' $r(i,k)$ reflects how well-suited $k$ is to serve $i$. These are messages are then updated iteratively. Once the messages converge (this is not guaranteed) then $\max_{k \in X} a(i,k) + r(i,k)$ will identify the cluster center for $i$, and these decisions will be consistent \cite{Frey}.

@@

There exist many other methods for clustering. Methods such as MeanShift and DBSCAN focus on finding clusters of high density \cite{Comaniciu, Ester}. We choose not to compare against density based methods since their objective differs widely from the center based methods above. In addition, we will focus on test data in which center-based methods seem appropriate. We also choose not to compare our proposed methods against Bayesian clustering methods such as Gaussian mixture models since these methods require many more parameters (representing the prior information). In addition, the $k$-means algorithm above can be thought of as a special case of the Gaussian mixture model when all the covariances are equal \cite{ESL}. Therefore, this algorithm can serve as a basis of comparison when both algorithms have the same available information.

@@

There are four choices of criteria for choosing which clusters to merge. 
In single linkage clustering (also called nearest neighbor), we measure the distance between two clusters to be the smallest pairwise distance and then merge the two closest clusters. The main potential drawback to this method is that two points within a cluster can be very far apart. Complete linkage clustering on the other hand measures the distance between two clusters to be the largest pairwise distance and then merges the two closest clusters. While points within a single cluster will be quite similar using this metric, points may not belong to their closest cluster. 
A method between these two extremes is group average linkage which looks at the average pairwise distance between clusters and merges the two closest clusters. This strikes a balance between the closeness and compactness of the two techniques above and gives more reasonable results in practice \cite{ESL}. Lastly, there is Ward's method which merges the two clusters so as to minimize the sum of each cluster's variance \cite{Ward}. All four of these methods can be implemented in the Lance-Williams recursive framework which updates the cluster distances efficiently \cite{Lance}.

@@

Again, several algorithms have been built upon agglomerative clustering to incorporate non-spherical cluster shapes and categorical data such as the CURE, ROCK, and Chameleon algorithms \cite{Guha_CURE, Guha_ROCK, Karypis} but do not relate as much to our proposed algorithm in objective. We also choose not compare our proposed method with Bayesian hierarchical methods for the same reasons as above.

@@

We will compare the above algorithms against $k$-means, $k$-medoids, PAM, affinity propagation, and Shi and Meila's spectral algorithm. All these methods except $k$-medoids and PAM are implemented in python's scikit-learn package. For comparisons against affinity propagation, we will first observe how many clusters the algorithm outputs and use that as input for our proposed algorithms. We implement the above local search algorithm for $p=2, 3,4$ so the computation time is reasonable.

Given a true clustering, we will use the adjusted mutual information (AMI) measure to score the accuracy of each method. AMI measures the agreement of two assignments and is normalized to reflect chance so that a perfect labeling receives a score of 1 and an independent random labeling receives a score of 0 in expectation \cite{Vinh}. For each data set, we will also report the $k$-median objective function value for that clustering, the guaranteed approximation ratio, and the computation time.

The data sets we chose to use for comparison were

NEED TO DECIDE:
OR library $p$-median.
